Simulating High-Dimensional Multivariate Data

### 2020-12-24 020-simulation-algorithm.Rmd
4th draft. Address Ed comments Part II

## NORTA in higher dimensions

Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
Even in low-dimensional spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@@

The *Mapping step* takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).

Our algorithm allows the generation high-dimensional data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall tau matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix. In particular, if we let X be MVN with N components and set

$S_P = \{ R_p : R_p is a correlation matrix for X \}$
$S_K = \{R_K : R_K is a Kendall tau matrix for a MVN \}$
$S_S = \{R_s : R_s is a Spearman correlation matrix for a MVN\}$

then there are 1-1 mappings between these sets.

So if a user of your package says "I want a 100 dimensional RV with NB marginals where R_s is the target Spearman correlation matrix", you can easily check whether R_S is an element of the "admissible set" S_S. The same if they say they want the same marginals but a Kendall matrix R_K. As is, you check (at least based on paper) whether the Spearman satisfies 2-D bounds which is ok for 2-D problems. So you can provide higher dimensional information about whether something can be done using your algorithm. Furthermore, I would guess that these sets of admissible S_K and S_S are not that restrictive - In particular R_p approx R_s for a MVN so that would suggest the admissible set S_S is not very restrictive because R_p can be any psd matrix with 1's on diagonal.

>>>>

*NORTA in higher dimensions*. Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
Even in low-dimensional spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@@

The "mapping step" takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).

Our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.

In particular, if we let $X$ be MVN with $d$ components and set

$$
\begin{aligned}
\Omega_P &= \{ R_P : R_P \,  is \, a \,  correlation \,  matrix \,  for \,  X \} \\
\Omega_K &= \{ R_K : R_K \,  is \,  a \,  Kendall \,  \tau  \, matrix \,  for \,  X \} \\
\Omega_S &= \{ R_S : R_S \,  is \,  a \,  Spearman \,  correlation \,  matrix \,  for \,  X\} \\
\end{aligned}
$$

then there are 1-1 mappings between these sets.
We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive.
In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PSD matrix with 1's along the diagonal. 
We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$) and also for checking that each pairwise correlation value is admissible (via the MC estimation of the 2-dimensional Frechet bounds).

Yet this scheme still suffers similar issues as the original Pearson-matching NORTA algorithm regarding the increasing probability of encountering a non-admissible correlation matrix as dimension increases.
In our experience, the *mapping step* for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix). The second step handles this issue by computing the nearest correlation matrix to $R_X$.
In the RNA-seq example (Section \@ref(examples)), we demonstrate this and show how to quantify and control the approximation error.
